Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-03-15 manuscript/manuscript.Rmd
minor typos fixed.

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in a recent and comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].

@@

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
df1<- df1 %>% 
  mutate_if(is.numeric, round, digits = 4) %>% 
  rename(Variables = X)
df1[1, 1] <- "width_90%"
df1[2, 1] <- "width_80%"
df1[3, 1] <- "width_70%"
df1[4, 1] <- "width_60%"
df1[5, 1] <- "width_50%"
df1[6, 1] <- "width_40%"
df1[7, 1] <- "width_30%"
df1[8, 1] <- "width_20%"
df1[9, 1] <- "width_10%"
df1[10, 1] <- "thickness_90%"
df1[11, 1] <- "thickness_80%"
df1[12, 1] <- "thickness_70%"
df1[13, 1] <- "thickness_60%"
df1[14, 1] <- "thickness_50%"
df1[15, 1] <- "thickness_40%"
df1[16, 1] <- "thickness_30%"
df1[17, 1] <- "thickness_20%"
df1[18, 1] <- "thickness_10%"
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components. PC1 (Dim.1) is postively correlated with all thickness-related variables and negatively correlated with all width-related variables and the maximum length. PC2 (Dim.2) is positively with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

The negative correlation between the PC1 and PC2 values suggests a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than half way across the surface while not breaking the handaxe into half [-@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182]. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the cultural reproduction of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

>>>>

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. At the same time, however, researchers have also pointed out that the technological choices defining a particular mental template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in a recent and comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].

@@

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
df1<- df1 %>% 
  mutate_if(is.numeric, round, digits = 4) %>% 
  rename(Variables = X)
df1[1, 1] <- "width_90%"
df1[2, 1] <- "width_80%"
df1[3, 1] <- "width_70%"
df1[4, 1] <- "width_60%"
df1[5, 1] <- "width_50%"
df1[6, 1] <- "width_40%"
df1[7, 1] <- "width_30%"
df1[8, 1] <- "width_20%"
df1[9, 1] <- "width_10%"
df1[10, 1] <- "thickness_90%"
df1[11, 1] <- "thickness_80%"
df1[12, 1] <- "thickness_70%"
df1[13, 1] <- "thickness_60%"
df1[14, 1] <- "thickness_50%"
df1[15, 1] <- "thickness_40%"
df1[16, 1] <- "thickness_30%"
df1[17, 1] <- "thickness_20%"
df1[18, 1] <- "thickness_10%"
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components. PC1 (Dim.1) is postively correlated with all thickness-related variables and negatively correlated with all width-related variables and the maximum length. PC2 (Dim.2) is positively correlated with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

The negative correlation between the PC1 and PC2 values suggests a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than half way across the surface while not breaking the handaxe into half [@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182]. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the cultural reproduction of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.
